Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2021-11-01 Presentations/ThirdYearPaperpres.tex

\begin{frame}{Technology and Displacement}

@@

\begin{frame}[fragile]{What is an Electronic Health Record?}
\begin{itemize}
    \item Technology purchased and implemented by hospitals
\end{itemize}

\begin{itemize}
    \item Used for:
    \vspace{3mm}
    \begin{itemize}
        \item Electronic patient records
        \vspace{3mm}
        \item Decision making assistance
    \end{itemize}
\end{itemize}

\note[item]{Get everyone on the same page about how to think about EHRs}
\note[item]{A product in the form of technology that is sold to hospitals (and others, but we focus on hospital setting)}
\note[item]{The most standard element of an EHR is that it converts paper records to electronic records}
\note[item]{New elements of EHRs: interropability, patient interface, decision-making assistance}

@@

\note[item]{Why do we care about EHRs or health information technology in the first place?}
\note[item]{Verbally explain the bridge to physicians here (the slide I took out)}
\end{frame}

\begin{frame}[noframenumbering]{Physician Response to EHRs}
\begin{center}
    \includegraphics[scale=.5]{graphics/News Clip2.PNG}
\end{center}
\end{frame}

\begin{frame}[noframenumbering]{Physician Response to EHRs}
\begin{center}
    \includegraphics[scale=.35]{graphics/News Clip3.PNG}
\end{center}
\end{frame}

@@

Did EHR implementation in hospitals affect physician labor market decisions?
\begin{itemize}
    \item Intensive Margin / Productivity
    \item Extensive Margin
    \item Hospital $\rightarrow$ Office
\end{itemize}
\end{frame}

@@

\begin{frame}{Literature}
\footnotesize 
\begin{columns}
    \column{0.34\textwidth}
        \centering
        \underline{ EHR $\rightarrow$ Healthcare }
        \begin{itemize}
            \item Small improvements for severe health conditions \\ \vspace{1mm}
            \tiny(Agha 2014; McCullough et al 2016)

@@

            \item No decrease in cost \\ \vspace{1mm}
            \tiny (Agha 2014; Dranove et al 2019)

@@

            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
            \tiny (Hitt $\&$ Tambe 2016, Butler $\&$ Johnson 2016, Meyerhoefer et al 2016)

@@

    \column{0.33\textwidth}
    \color{gray}
        \centering
        \underline{ Technology $\rightarrow$ Labor }
        \begin{itemize}
        \color{gray}
            \item Erosion effect vs. wage effect \\ \vspace{1mm}
            \tiny (Zeira $\&$ Joseph 2011)

            \item Cause retirement? \\ \vspace{1mm}
            \tiny (Schleife 2006, Cavapozzi et al 2013, Roger et al 2006, Friedburg 2003)

            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
            \tiny (Willis and Hudomiet 2020)
        \end{itemize}

    \column{0.33\textwidth}
        \color{gray}
        \centering
        Physician Labor \underline{ Determinants }
        \begin{itemize}
        \color{gray}
            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
            \tiny (Baltagi et al 2005, Sloan 2018)

            \item Peronal life factors matter more for retirement than work conditions \\ \vspace{1mm}
            \tiny (Bahrami 2002)

            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
            \tiny (Dewa et al 2014)
        \end{itemize}
\end{columns}

\note[item]{\footnotesize Agha 2014 (JHE) uses claims data and difference in difference methodology. Find 1.3 percent increase in billed costs and no improvements in health coutcomes, data is 1998-2005.}

\note[item]{McCullough, Parente, Town 2016, RAND. claims data 2002-2007. Find no reduction in moretality for the median patient, only those who require complex care with coordination.}

\note[item]{Appari, Johnson, Anthony 2013. Health services review. fixed effects linear panel model. quality measure comes from hospital compare, not at the patient level.}

\note[item]{Dranove, Forman, Goldfarb, Greenstein (2019) American Economic Journal: Economic Policy. data from 1996 to 2009. Specific findings: hospitals have a rise in cost after 3 years, some hospitals in certain locations have a decrease in costs after 5 years but others still have a higher cost after 6 years.}

@@

\begin{frame}[noframenumbering]{Literature}
\footnotesize 
\begin{columns}
    \column{0.34\textwidth}
    \color{gray}
        \centering
        \underline{ EHR $\rightarrow$ Healthcare }
        \begin{itemize}
        \color{gray}
            \item Small improvements for severe health conditions \\ \vspace{1mm}
            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND; Meyerhoefer et al, ILRR)

            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
            \tiny (Appari et al 2013)

            \item No decrease in cost \\ \vspace{1mm}
            \tiny (Agha 2014, JHE; Dranove et al 2019; AEJ)

            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
            \tiny (Hitt $\&$ Tambe 2016, Butler $\&$ Johnson 2016, Meyerhoefer et al 2016)

    \column{0.33\textwidth}
        \centering
        \underline{ Technology $\rightarrow$ Labor }
        \begin{itemize}
            \item Erosion effect vs. wage effect \\ \vspace{1mm}
            \tiny (Zeira $\&$ Joseph 2011)

            \item Cause retirement? \\ \vspace{1mm}
            \tiny (Schleife 2006, Cavapozzi et al 2013, Roger et al 2006, Friedburg 2003)

            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
            \tiny (Willis and Hudomiet 2020)
        \end{itemize}

    \column{0.33\textwidth}
    \color{gray}
        \centering
        Physician Labor \underline{ Determinants }
        \begin{itemize}
        \color{gray}
            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
            \tiny (Baltagi et al 2005, Sloan 2018)

            \item Peronal life factors matter more for retirement than work conditions \\ \vspace{1mm}
            \tiny (Bahrami 2002)

            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
            \tiny (Dewa et al 2014)
        \end{itemize}
\end{columns}

\note[item]{Zeira and Johnson 2011. }
\end{frame}

\begin{frame}[noframenumbering]{Literature}
\footnotesize 
\begin{columns}
    \column{0.34\textwidth}
    \color{gray}
        \centering
        \underline{ EHR $\rightarrow$ Healthcare }
        \begin{itemize}
        \color{gray}
            \item Small improvements for severe health conditions \\ \vspace{1mm}
            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND; Meyerhoefer et al, ILRR)

            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
            \tiny (Appari et al 2013)

            \item No decrease in cost \\ \vspace{1mm}
            \tiny (Agha 2014, JHE; Dranove et al 2019; AEJ)

            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
            \tiny (Hitt $\&$ Tambe 2016, Butler $\&$ Johnson 2016, Meyerhoefer et al 2016)

    \column{0.33\textwidth}
    \color{gray}
        \centering
        \underline{ Technology $\rightarrow$ Labor }
        \begin{itemize}
        \color{gray}
            \item Erosion effect vs. wage effect \\ \vspace{1mm}
            \tiny (Zeira $\&$ Joseph 2011)

            \item Cause retirement? \\ \vspace{1mm}
            \tiny (Schleife 2006, Cavapozzi et al 2013, Roger et al 2006, Friedburg 2003)

            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
            \tiny (Willis and Hudomiet 2020)
        \end{itemize}

    \column{0.33\textwidth}
        \centering
        Physician Labor \underline{ Determinants }
        \begin{itemize}
            \item Physicians not very responsive to wage changes in terms of hours worked \\ \vspace{1mm}
            \tiny (Baltagi et al 2005, Sloan 2018)

            \item Peronal life factors matter more for retirement than work conditions \\ \vspace{1mm}
            \tiny (Bahrami 2002)

            \item Physician burnout: early retirement and less hours worked \\ \vspace{1mm} 
            \tiny (Dewa et al 2014)
        \end{itemize}
\end{columns}
\end{frame}

@@

\begin{frame}{About the Data}
    Main data: \underline{CMS Shared Patient Data} (2009-2015)
    \begin{itemize}
        \item Lists 2 NPIs and the number of billings those two entities shared
    \end{itemize}
    \vspace{3mm}

    How I use it:
    \begin{itemize}
        \item Physician-hospital pairs (PCPs, internists, hospitalists)
        \item Billings that occur on same day
        \item Close relationship between hospital and physician
        \item Aggregate to the physician level
    \end{itemize}
\end{frame}

\begin{frame}{Measures of Physician Labor Market Decisions}

     Patients visited in hospitals

     Indicator for switching work setting:
    \begin{itemize}
        \item 0 patients visited but maintaining billable activity
    \end{itemize}

    \textcolor{gray}
     {Indicator for retirement/ leaving the labor force}

@@

\begin{frame}{Measure of EHR Use}

Physician-level treatment variable that captures exposure to \textit{any} EHR
\vspace{3mm}
\begin{itemize}
    \item Individual hospital EHR use comes from AHA survey
    \vspace{3mm}
    \item Treatment turns on the first time any of the physicians' hospitals implements an EHR
\end{itemize}
\end{frame}

\begin{frame}{Other Characteristics}

Medical school graduation date: \underline{Physician Compare}

@@

\begin{frame}{Data}
\centering
    \includegraphics[scale=.8]{Objects/sumstats.pdf}

\begin{frame}{Physician EHR Use}
    \centering
    \includegraphics[scale=.5]{Objects/sum_stats_year.pdf}
\end{frame}

@@

@@

\begin{frame}{Dep. Variable: Patients Billed with Hospitals}
\begin{figure}[ht]
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_fullsample.pdf}
            \caption{\small All Physicians \\(Only 1 Hospital)}
        \end{minipage}
        \hspace{0.2cm}
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_oldsample.pdf}
            \caption{\small Phys. $>$ 35yrs\\ (Only 1 Hospital)}
        \end{minipage}
    \end{figure}
\end{frame}

\begin{frame}{Dep. Variable: Probability of Billing, but not With Hospitals}
\begin{figure}[ht]
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/xtevent_fullsample_ind.pdf}
            \caption{\small All Physicians}
        \end{minipage}
        \hspace{0.2cm}
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/xtevent_oldsample_ind.pdf}
            \caption{\small Phys. $>$ 35yrs Experience}
        \end{minipage}
    \end{figure}
\end{frame}

\begin{frame}{Dep. Variable: Probability of Billing, but not With Hospitals}
\begin{figure}[ht]
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_fullsample_ind.pdf}
            \caption{\small All Physicians \\(Only 1 Hospital)}
        \end{minipage}
        \hspace{0.2cm}
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_oldsample_ind.pdf}
            \caption{\small Phys. $>$ 35yrs Experience\\ (Only 1 Hospital)}
        \end{minipage}
    \end{figure}
\end{frame}

\begin{frame}{Dep. Variable: Appears in Office-Based Data}
\begin{figure}[ht]
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/xtevent_fullsample_office.pdf}
            \caption{\small All Physicians}
        \end{minipage}
        \hspace{0.2cm}
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/xtevent_oldsample_office.pdf}
            \caption{\small Phys.$>$35yrs Experience}
        \end{minipage}
    \end{figure}
\end{frame}

\begin{frame}{Dep. Variable: Appears in Office-Based Data}
\begin{figure}[ht]
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_fullsample_office.pdf}
            \caption{\small All Physicians \\(Only 1 Hospital)}
        \end{minipage}
        \hspace{0.2cm}
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/xtevent_hosp_oldsample_office.pdf}
            \caption{\small Phys.$>$35yrs Experience\\ (Only 1 Hospital)}
        \end{minipage}
    \end{figure}
\end{frame}

>>>>

\begin{frame}{The Broad Question}
\begin{itemize}
    \item Longstanding topic in economics: relationship between technology and employment
    \vspace{2mm}
    \begin{itemize}
        \item Displacement?
        \item Productivity?
        \item Wages?
    \end{itemize}
    \vspace{3mm}
    \item Investigated in many settings
    \vspace{3mm}
    \item This paper: specific healthcare setting
\end{itemize}

@@

\begin{frame}[fragile]{What is an Electronic Health Record?}
``An electronic health record is the systematized collection of patient and population electronically stored health information in a digital format"

\vspace{3mm}

What it can do:
\begin{itemize}
    \item Paper $\rightarrow$ computerization
    \item Communication
    \item Decision making assistance
\end{itemize}

@@

\begin{frame}{Physician Response to EHRs}
\begin{center}
    \includegraphics[scale=.3]{graphics/News Clip3.PNG}

    \vspace{3mm}

    \includegraphics[scale=.3]{graphics/News Clip2.PNG}
\end{center}

@@

Did EHR implementation in hospitals affect physician labor market decisions?
\begin{itemize}
    \item Intensive Margin / Productivity
    \item Extensive Margin
    \item Hospital $\rightarrow$ Office
\end{itemize}

@@

\begin{frame}{Literature}
\small
\begin{columns}
\setlength{\tabcolsep}{-5pt}
    \column{0.4\textwidth}
        \centering
        \underline{ Technology $\rightarrow$ Labor }
        \vspace{-1mm}
        \begin{itemize}
            \item Erosion effect vs. wage effect \\ \vspace{1mm}
            \tiny (Zeira $\&$ Joseph 2011, EJ)

@@

            \item Older, well educated members of labor force exited at higher rates due to computerization \\ \vspace{1mm}
            \tiny (Willis and Hudomiet 2021, Labour)
        \end{itemize}

        \column{0.4\textwidth}
        \centering
        \color{gray}
        \underline{ EHR $\rightarrow$ Healthcare }
        \vspace{-1mm}
        \begin{itemize}
        \color{gray}
            \item Small improvements for severe health conditions \\ \vspace{1mm}
            \tiny(Agha 2014, JHE; McCullough et al 2016, RAND)

@@

            \item Higher ``levels" of EHR use $\rightarrow$ decline in quality of care\\ \vspace{1mm}
            \tiny (Appari et al 2013, HSR)

            \item No decrease in cost \\ \vspace{1mm}
            \tiny (Agha 2014, JHE; Dranove et al 2019, AEJ)

            \item Productivity and opinion about EHRs varies drastically \\ \vspace{1mm}
            \tiny (Hitt $\&$ Tambe 2016, ILRR; Butler $\&$ Johnson 2016, IJHEM; Meyerhoefer et al 2016, ILRR)

@@

\end{columns}

@@

@@

\begin{frame}{Data}
    \underline{Main: CMS Shared Patient Data (2009-2015)}
    \begin{itemize}
        \item Hospital-physician pairs: bill for same patient under Medicare
        \item Physician (PCP, hospitalist, internist) works in the hospital
    \end{itemize}

    \vspace{3mm}

    \underline{EHR Use:}
    \begin{itemize}
        \item AHA Survey
    \end{itemize}

    \vspace{3mm}

    \underline{Other:}
    \begin{itemize}
        \item Physician Compare
    \end{itemize}

    \vspace{3mm}

    Aggregated to physician level
\end{frame}

\begin{frame}{Variables}
Labor Market Outcomes:
\vspace{-2mm}
\begin{itemize}
    \item Patients visited in hospitals
    \item \color{gray} Maintaining billable activity only outside of hospitals
    \item \color{gray} Level of billable activity outside of hospitals
    \item \color{gray} Retirement/leaving the labor force
\end{itemize}

EHR Use:
\vspace{-2mm}
\begin{itemize}
    \item Indicator for any exposure to EHR
    \item \color{gray} Fraction of hospitals exposed to EHR
\end{itemize}

Other:
\vspace{-2mm}
\begin{itemize}
    \item Years of experience
    \item Hospital characteristics
\end{itemize}

@@

\begin{frame}{Data}
    \centering
    \includegraphics[scale=.5]{Objects/sum_stats_year.pdf}
\end{frame}

\begin{frame}{Data}
\centering
    \includegraphics[scale=.5]{Objects/cont_treatment_graph.pdf}

@@

@@

\begin{align*}
    y_{it} &: \text{Outcomes} \hspace{30mm}\\
    \alpha_i, \delta_t &: \text{Fixed Effects}\hspace{30mm}\\
    q_{it} &: \text{Physician/hospital characteristics}\hspace{30mm}\\
    z_{i,t-k} &: \text{Exposure to EHR} \hspace{30mm}
\end{align*}

@@
